---
title: "Download statistics from University of Cambridge for ScienceDirect"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Introduction

Through an FOI request, I have obtained the download statistics
(COUNTER 5 format) for the calendar years 2019 and 2020 for
ScienceDirect Usage at the University of Cambridge.  These data
exclude any Gold OA articles.

The raw data are available in the [data](https://github.com/sje30/counter5/tree/main/data/) folder.  Below is some
simple analysis of the data; please let me know if you see any errors
in the analysis, particularly in the handling of the COUNTER R5 data.
I have chosen to analyse the 'Total_Item_Requests', rather than the 
'Unique_Item_Requests'.

For each of the two years I present a searchable list of
downloads/year that is ranked in decreasing order of downloads.  These
data are then converted into a cumulative histogram of access.

I have made no attempt currently to break down the journals into
different groupings (e.g. Freedom collection).  This file is available
as a Rmarkdown file.



```{r init, echo=FALSE, results=FALSE, message=FALSE}
require(tidyverse)
require(readr)
require(dplyr)

read_data <- function(file) {
  ## Return the journals and number of downloads per year.
  ## Skip over the header of the file
  dat <- read_csv(file, skip=13)

  ## Ignore the YOP = 9999 for now (are they in the future?)
  d2 = filter(dat, Metric_Type =="Total_Item_Requests", YOP!=9999 ) %>%
    select(Title, YOP, Reporting_Period_Total)

  summary2 <- d2 %>%
    group_by(Title) %>%
    summarise(requests = sum(Reporting_Period_Total), n = n()) %>%
    arrange(desc(requests)) %>%
    mutate(rank=1:n())

  ## get the cumulative percentage of downloads by this journal
  total_requests = sum(summary2$requests)
  pct = 100.0 * round(summary2$requests / total_requests, 3)
  c_pct = 100.0 * round(cumsum(summary2$requests) / total_requests, 3)
  summary2 <- summary2 %>% mutate(pct=pct, c_pct=c_pct)

  summary2
}

cumulative_plot <- function(tbl, main=NULL) {
  plot(tbl$c_pct, main=main,
       xlab='Journals', bty='n', las=1, ylim=c(0,100),
       type='l', ylab='cumulative percentage of downloads', lwd=2)
  abline(h=80, col='blue')
  x = 0.2 * nrow(tbl)
  abline(v=x, col='blue')
}
```


# 2019 data

```{r echo=FALSE, message=FALSE,results=FALSE}
dat_2019 <- read_data("data/ScienceDirect_TR_J4_Standard_R5_Jan-2019_Dec-2019_University+of+Cambridge_20211014_1631.csv")
```

Total number of downloads: 
`r format(sum(dat_2019$requests), scientific=FALSE)` from 
`r nrow(dat_2019)` journals.


Note: these tables are searchable; just type in e.g. a journal name in
the search box in the top right and it will narrow down the selection;
likewise, you can sort by other columns by clicking on an arrow next
to the column name.

The columns in the table are:

- Title: name of the journal
- requests: yearly total number of requests from that title
- n: number of distinct rows in the CSV for this title.
- rank: rank of the requests (highest first).
- pct: requests for this journal as a fraction of all journals.
- c_pct: cumulative version of the percentage, following the ranking.

```{r echo=FALSE}
DT::datatable(dat_2019,
              rownames=FALSE,
             caption="Download statistics by journal, 2019.")
```



# 2020 data

```{r echo=FALSE, message=FALSE,results=FALSE}
dat_2020 <- read_data("data/ScienceDirect_TR_J4_Standard_R5_Jan-2020_Dec-2020_University+of+Cambridge_20211014_1708.csv")
```

Total number of downloads: 
`r format(sum(dat_2020$requests), scientific=FALSE)` from 
`r nrow(dat_2020)` journals.

```{r echo=FALSE}
DT::datatable(dat_2020,
              rownames=FALSE,
             caption="Download statistics by journal, 2020.")
```

# Cumulative plots

These plots for 2019 and 2020 indicate that the data seem to follow
the [Pareto principle](https://en.wikipedia.org/wiki/Pareto_principle): 20% of
journals in the collection account for about 80% of downloads.

```{r,echo=FALSE,fig.cap='Cumulative Downloads for 2019 and 2020',fig.width=12}
par(mfrow=c(1,2))
cumulative_plot(dat_2019, main='2019 Downloads')
cumulative_plot(dat_2020, main='2020 Downloads')
```


# Discussion

It is unsurprising to see that the journals that are most accessed are
those that come from Cell Press and Lancet titles. The top-ranked
journal, Cell, alone counts for 3-4% of all downloads.

Statistics like this are interesting to view, but should be viewed
with caution (e.g. Wood-Doughty et al 2019).  I believe they are
useful to help inform the value of a big deal versus unbundling
(Thornton and Brundy 2021).  However, over-reliance on simple metrics
like this to select which journals to keep and which to remove may
disproportionality affect smaller disciplines.  As soon as these
metrics drive decision making may lead to 'gaming', whereby
researchers routinely download papers from their favourite journals
simply to prevent them being cancelled.  Further, I'd hope that such
statistics may not even be relevant in a few years, if we can
transition to more equitable models of publishing.


## Questions / future work

- Why are there so many journals (about 500) with only one download in
  the year?  Why are there none with zero downloads?
  
- For curiosity, what is the distribution of downloads over the year
  (I've ignored monthly totals for now)?  Is there a seasonal effect
  to download statistics?

- Is it worth collecting this COUNTER R5 data from other UK
  institutions?  Should it be freely available as a matter of routine?

- Can this be combined with costs of titles to evaluate the cost of
  subsets of journals?

See the [project home page](https://github.com/sje30/counter5) for all
source material and e.g. any github issues.


# References

Thornton JB, Brundy C (2021) Elsevier title level pricing: Dissecting
the bowl of spaghetti. J Libr Sch Commun 9:2410 Available at:
http://dx.doi.org/10.7710/2162-3309.2410.

Wood-Doughty A, Bergstrom T, Steigerwald DG (2019) Do Download Reports
Reliably Measure Journal Usage? Trusting the Fox to Count Your Hens?
Coll Res Libr 80:694 Available at:
https://crl.acrl.org/index.php/crl/article/view/17824/19653 [Accessed
October 16, 2021].

